A finite-state morphological transducer for Kyrgyz

### 2011-10-17 dev/paper/paper.tex
more clarifications
addition of LREC style files and some bib updates
a few fixes and updates

\documentclass[a4paper,12pt,onecolumn,oneside]{article}
\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}

@@

\usepackage[novoc,fdf2alif]{arabxetex}
\newfontfamily\uighurfont[Script=Arabic,Scale=1.5]{Lateef}

@@

\section{Introduction}
This paper describes the development of a morphological transducer oriented for the task of machine translation for the Kyrgyz language using the free/open-source platform HFST. The transducer was developed under the auspices of the Apertium \citet{forcada2011} project for use in a machine translation system from Turkish to Kyrgyz.

@@

Kyrgyz is spoken mostly in Kyrgyzstan where it has official status as the national language.  Many Kyrgyz speakers in Kyrgyzstan are bilingual in Russian and/or Uzbek, and make up a majority of the population of the country.  There are other sizable Kyrgyz-speaking communities outside of Kyrgyzstan, most notably in China (where the Kyrgyz are an officially recognised minority), Tajikistan, and Uzbekistan.  Current estimates of the number of speakers range from 3 million to 4 million.\footnote{Based on figures from \citet{lewis2009} and \citet{factbook2009}}  Not all ethnic Kyrgyz speak the language, and not all competent speakers are ethnic Kyrgyz, but there is a very strong correspondence between ethnic identity and knowledge of the language.\footnote{This interpretation of the situation is supported by the experiences of the authors with the language, and is common knowledge in Kyrgyzstan.}

%FIXME: Should we say something about how I sometimes consulted Tolgonay, or about where my knowledge of Kyrgyz comes from in the first place?
The understanding of Kyrgyz grammar employed to construct the transducer for this project was gained from the Kyrgyz knowledge of two of our authors.  Mirlan Ipasov is a native speaker of Kyrgyz, and Jonathan Washington is a theoretical and descriptive linguist fluent in Kyrgyz and knowledgeable about Turkic languages.  Some grammar sources were consulted, such as \citet{hebertpoppe1963}, \citet{usonalievomuraliev2003}, \citet{qudaybergenov1980}, \citet{somfaikara2003}, and \citet{imart1981}, but they were largely not relied on due to their approaches to Kyrgyz grammar.  Some dictionaries were also consulted, including \citet{jumakunova2005} and \citetwoyears{yudakhin1957}{yudakhin1965}.

@@

There are two rules that can immediately deal with some of these forms: optional loss of /D/ in the ablative suffix after 1st person singular, 2nd person singular, and 3rd person possession suffixes, and mandatory loss of /G/ in the dative suffix in the same situations.  Since these are rules specific to these morphological forms and don’t apply generally in Kyrgyz at a phonological level, they were implemented directly in the interaction of the continuation lexica for these possessive suffixes and ablative and dative case suffixes.
%FIXME: should we be more specific as to how we did this, or is something very vague and general like this good?

@@

\bibliographystyle{apa}
\bibliography{paper}

>>>>

\documentclass[10pt, a4paper]{article}

\usepackage{lrec2006}

\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}

@@

\usepackage[novoc,fdf2alif]{arabxetex}
\newfontfamily\uighurfont[Script=Arabic,Scale=1.5]{Lateef}
\newfontfamily\arabicfont[Script=Arabic,Scale=1.5]{Lateef}

@@

\section{Introduction}
This paper describes the development of a morphological transducer oriented for the task of machine translation for the Kyrgyz language using the free/open-source platform HFST. The transducer was developed under the auspices of the Apertium \cite{forcada2011} project for use in a machine translation system from Turkish to Kyrgyz.

@@

Kyrgyz is spoken mostly in Kyrgyzstan where it has official status as the national language.  Many Kyrgyz speakers in Kyrgyzstan are bilingual in Russian and/or Uzbek, and make up a majority of the population of the country.  There are other sizable Kyrgyz-speaking communities outside of Kyrgyzstan, most notably in China (where the Kyrgyz are an officially recognised minority), Tajikistan, and Uzbekistan.  Current estimates of the number of speakers range from 3 million to 4 million.\footnote{Based on figures from \cite{lewis2009} and \cite{factbook2009}}  Not all ethnic Kyrgyz speak the language, and not all competent speakers are ethnic Kyrgyz, but there is a very strong correspondence between ethnic identity and knowledge of the language.\footnote{This interpretation of the situation is supported by the experiences of the authors with the language, and is common knowledge in Kyrgyzstan.}

%FIXME: Should we say something about how I sometimes consulted Tolgonay, or about where my knowledge of Kyrgyz comes from in the first place?
The understanding of Kyrgyz grammar employed to construct the transducer for this project was gained from the Kyrgyz knowledge of two of our authors.  Mirlan Ipasov is a native speaker of Kyrgyz, and Jonathan Washington is a theoretical and descriptive linguist fluent in Kyrgyz and knowledgeable about Turkic languages.  Some grammar sources were consulted, such as \cite{hebertpoppe1963}, \cite{usonalievomuraliev2003}, \cite{qudaybergenov1980}, \cite{somfaikara2003}, and \cite{imart1981}, but they were largely not relied on due to their approaches to Kyrgyz grammar.  Some dictionaries were also consulted, including \cite{jumakunova2005} and \citetwoyears{yudakhin1957}{yudakhin1965}.

@@

%FIXME: maybe a footnote at least about what "optional loss of /D/" means?
%TODO: some explanation of how optional rules were dealt with
There are two rules that can immediately deal with some of these forms: optional loss of /D/ in the ablative suffix after 1st person singular, 2nd person singular, and 3rd person possession suffixes, and mandatory loss of /G/ in the dative suffix in the same situations.  Since these are rules specific to these morphological forms and don’t apply generally in Kyrgyz at a phonological level, they were implemented directly in the interaction of the continuation lexica for these possessive suffixes and ablative and dative case suffixes.
%FIXME: should we be more specific as to how we did this, or is something very vague and general like this good?

@@

\bibliographystyle{lrec2006}
\bibliography{paper}
